---
title: "U.S. Egg Production Analysis Using Linear Mixed Models"
subtitle: "An Exploration Across States and Months"
author: "Rose Dalton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
  pdf:
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{tabularx}
course: Capstone Projects in Data Science
bibliography: references.bib 
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

Statistical methods for comparing differences between groups have long
served as a cornerstone of empirical research across the natural and
social sciences. Foundational tools such as t-tests, analysis of
variance (ANOVA), and linear regression provided early mechanisms for
evaluating mean differences and variance relationships under specific
conditions. However, as research designs grew more complex,
incorporating repeated measurements, nested data structures, and
unbalanced datasets, the limitations of traditional techniques became
increasingly apparent. These methodological pressures contributed to the
development of linear mixed-effects models (LMMs), which extend
classical linear modeling by incorporating both fixed and random sources
of variation within hierarchical data structures. The purpose of this
review is to examine how LMMs came to prominent usage, the associated
benefits and limitations, and how this method may be applied to this
capstone project.

Formal statistical tests for group comparisons experienced widespread
growth in the early twentieth century. Sir Ronald Fisher’s Statistical
Methods for Research Workers (1925) was one of the earliest compilations
of comparison analyses such as t-tests, the analysis of variance
(ANOVA), the maximum likelihood estimation, and experimental design for
correlation studies (@huitema2025fisher). Building on this groundwork,
Henderson (1953) advanced methods for estimating variance and covariance
components in unbalanced agricultural datasets. His work demonstrated
how linear models could utilize variance elements from both fixed and
random factors, providing unbiased estimates even among unbalanced group
structures. These contributions laid critical groundwork for the
development of modern mixed-effects modeling.

Traditional approaches such as ANOVA and simple regression perform
optimally under assumptions of independence, normality, and balanced
groupings (@henderson1953variance). Violations to the general schema —
including repeated measures, nested clustering, and missing data — can
compromise inference (@brown2021introduction). Linear mixed-effects
models address these limitations by modeling fixed population-level
effects alongside random group-level deviations within a single
framework. This hierarchical variance structure allows LMMs to account
for correlated observations while improving estimate stability through
partial pooling, where group estimates are shrunk toward the overall
mean when data are limited or missing(@meteyard2020best).

This structure endows LMMs with considerable flexibility.
@schielzeth2020robustness demonstrated that fixed-effect variance
estimates remain largely unbiased under assumption violations such as
skewness, heteroskedasticity, and bimodality, though precision may
decline. Despite these advantages, LMMs introduce analytical
complexities. Model specification — particularly the selection of
random-effects representation terms— can be challenging, and over- or
under-parameterization may affect model stability and interpretability
(@ryoo2011model). Additionally, while LMMs accommodate unbalanced data,
reduced precision under distributional violations remains an important
consideration (@schielzeth2020robustness).

Owing to their robust performance in the face of imperfect data, LMMs
are commonly utilized in practical fields such as agriculture,
psychology, ecology, and clinical research. @meteyard2020best documented
their extensive adoption in psychological research, while agricultural
applications include modeling spatial yield variation in plant breeding
trials (@adhikari2016comparing) and assessing environmental drivers of
winter wheat performance (@zhou2022using).

Within the present capstone project, LMMs provide an appropriate
framework for modeling U.S. egg production across states and years while
investigating the effects of average feed price and temperature. The
repeated-measures and nested structure of the dataset necessitate an
approach capable of partitioning population-level effects from
state-level variability. By capturing both fixed interactions and
hierarchical random variation, LMMs offer more comprehensive insights
into production dynamics than traditional regression alone.

## Methods

The general linear mixed model may be applied to the egg production
dataset to examine factors associated with laying productivity. Linear
mixed models are commonly used to analyze hierarchical or longitudinal
datasets in which repeated observations are nested within higher-level
groups(@meteyard2020best). The response variable, $\boldsymbol{y}$,
represents eggs produced per 100 laying hens, a standard measure of
laying productivity reported in USDA egg production statistics. Feed
price, yearly variation, monthly variation, and average monthly
temperature by state are represented in the fixed-effects portion of the
model, $\boldsymbol{\beta}$. Fixed effects are used to estimate the
systematic contribution of these variables to egg productivity across
the population of commercial egg-producing states. In contrast,
inter-state variation is captured through the random-effects component
$\boldsymbol{\gamma}$, while the residual error term $\mathbf{e}$
represents fluctuations in productivity not explained by either the
fixed or random components of the
model(@Mustafa2023_FixedRandomEffects_Medium).

In this capstone project, the fixed-effects portion of the model
represents systematic relationships between egg productivity and
explanatory variables that are assumed to influence all states in a
similar manner(@Mustafa2023_FixedRandomEffects_Medium). Average monthly
temperature and feed price are included as fixed effects because they
represent measurable environmental and economic conditions expected to
affect egg production across the industry as a whole. The associated
fixed-effect coefficients therefore estimate the average change in eggs
produced per 100 layers associated with variation in these predictors,
while also accounting for seasonal and long-term patterns through month
and year effects.

The random-effects component accounts for unobserved heterogeneity
between states. Commercial egg production systems vary across states due
to differences in housing systems, farm size, management practices,
breed composition, and regional climate conditions(@incredibleegg2013).
These factors may lead to systematic differences in baseline
productivity that are not fully captured by the measured predictors. To
accommodate this variability, the model includes state-level random
effects, allowing each state to deviate from the overall intercept. This
structure reflects the hierarchical nature of the dataset, in which
repeated monthly observations are nested within states.

In this study, state is modeled as a random effect rather than a fixed
effect because the goal is not to estimate a separate parameter for each
individual state, but rather to account for variability across the
broader population of egg-producing states. Treating state as a random
effect allows each state to have its own baseline productivity level
while still estimating overall relationships between temperature, feed
price, month, year, and egg production. This approach is appropriate
when observations are grouped and repeatedly measured within the same
units over time(@sarafian2020lmm). Additionally, the modeling of state
as a random effect reduces the number of parameters required compared
with estimating separate fixed coefficients for each state and improves
the model’s ability to generalize the estimated relationships beyond the
specific states included in the
dataset(@Mustafa2023_FixedRandomEffects_Medium).

------------------------------------------------------------------------

The general linear mixed model (LMM) is defined below:

$$
\mathbf{y} = X\boldsymbol{\beta} + Z\boldsymbol{\gamma} + \mathbf{e}
$$

-   $\mathbf{y}$: outcome vector

-   $X$: fixed-effects matrix

-   $\boldsymbol{\beta}$: fixed-effects coefficient vector

-   $Z$: random-effects matrix

-   $\boldsymbol{\gamma}$: random-effects vector

-   $\mathbf{e}$: residual error vector

------------------------------------------------------------------------

The residual error term satisfies:

$$
\mathbf{e} \sim \mathcal{N}(\mathbf{0}, R)
$$

-   $\boldsymbol{0}$: mean

-   ${R}$: variance

------------------------------------------------------------------------

The random effects vector satisfies:

$$
\boldsymbol{\gamma} \sim \mathcal{N}(\mathbf{0}, G)
$$

-   $\boldsymbol{0}$: mean

-   ${G}$: variance-covariance matrix

------------------------------------------------------------------------

Additionally, the residual errors and random effects are assumed
uncorrelated(@Ao2007_MixedModels_HBS):

$$
\mathrm{Cov}(\mathbf{e}, \boldsymbol{\gamma}) = 0
$$

## Analysis and Results

### Data Exploration and Visualization

The dataset was compiled from multiple government agency resources. Egg
production rates and grain prices were sourced from the U.S. Department
of Agriculture’s National Agricultural Statistics Service monthly report
archive (@USDA_ChickensAndEggs_2026, @USDA_AgriculturalPrices_2026).
Temperature data was sourced from the National Oceanic and Atmospheric
Administration’s Climate at a Glance National Time Series archive
(@NOAA_ClimateAtAGlance_2026).

| Variable | Type | Description |
|------------------------|------------------------|------------------------|
| `date` | Time Index | Monthly observation period from January 2000 through December 2024. |
| `state` | Categorical | 25 U.S. states with consistently high commercial egg production. |
| `eggsPer100Layers` | Continuous | Monthly productivity measure defined as the number of eggs produced per 100 laying hens. |
| `cornPrice` | Continuous | Monthly average U.S. price per bushel of corn (USD). |
| `soybeanPrice` | Continuous | Monthly average U.S. price per bushel of soybeans (USD). |
| `feedPrice` | Continuous | Weighted feed cost calculated as 2/3 × `cornPrice` + 1/3 × `soybeanPrice` (USD per bushel). |
| `avgTemp` | Continuous | State-level monthly average temperature (°F). |

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages
library(tidyverse)
library(ggthemes)

```

```{r, warning=FALSE, echo=TRUE}
# load data
eggData <- read.csv("USDA_Egg_Data.csv")

# clean data for visuals
eggData_clean <- eggData %>%
  select(date, state, eggsPer100Layers, avgTemp, feedPrice) %>%
  mutate(
    state = str_trim(state),                         
    eggsPer100Layers = as.numeric(eggsPer100Layers), 
    avgTemp = as.numeric(avgTemp),
    Date = as.Date(paste0(date, "-01")),             
    Year = year(Date),                              
    Month = month(Date)
  ) 

#time series plot
timeSeries = eggData_clean %>% ggplot(mapping = aes(x=Date, y=eggsPer100Layers)) 

timeSeries + geom_line(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, span = 0.25) +
  facet_wrap(~ state, ncol = 5, scales = "free_y") +
  labs(title = "Eggs per 100 Layers Over Time (2000–2024)",
    x = "Date",
    y = "Eggs per 100 Layers",
    caption = "Figure 1") +
  theme_bw()
```

Seasonality appears present for every state. This is evidenced by the regularly repeating peaks and troughs. We can presume that laying productivity follows an annual pattern. While nearly all states examined exhibit some degree of increase in egg productivity, Alabama yielded a unique graph. This trend line is generally consistent over the 25 years with minor variability present.

```{r, warning=FALSE, echo=TRUE}

# month box plot
boxPlot = eggData_clean %>% ggplot(mapping = aes(x=factor(Month), y=eggsPer100Layers))

boxPlot + geom_boxplot(outlier.alpha = 0.3) +
  labs(title = "Seasonality: Eggs per 100 Layers by Month",
    x = "Month",
    y = "Eggs per 100 Layers",
    caption = "Figure 2") +
  theme_bw()
```

This plot indicates that there are disparities in egg production across months. Productivity appears to be higher in January, March, May, July, August, October, and December. The least productive month is evidently February.

```{r, warning=FALSE, echo=TRUE}

# temperature scatterplot
scatter = eggData_clean %>% ggplot(mapping = aes(x=avgTemp, y=eggsPer100Layers))

scatter + geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75) +
  labs(title = "Eggs per 100 Layers vs Average Temperature",
    x = "Average Temperature (°F)",
    y = "Eggs per 100 Layers",
    caption = "Figure 3") +
  theme_bw()
```

A notable feature of this scatterplot is how the productivity trend declines sharply as temperature exceeds 75 degrees F. @Kim2024_HeatStress_LayingHens explains that laying hens are susceptible to heat stress under high temperature conditions. A temperature–humidity index of 76-81 degrees is considered dangerous and over 81 is categorized as emergency.

```{r, warning=FALSE, echo=TRUE}

# feed scatterplot
scatter2 = eggData_clean %>% ggplot(mapping = aes(x=feedPrice, y=eggsPer100Layers))

scatter2 + geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75) +
  labs(title = "Eggs per 100 Layers vs Feed Price",
       x = "Feed Price (USD/bushel)",
       y = "Eggs per 100 Layers",
    caption = "Figure 4") +
  theme_bw()
```

This scatterplot seems to indicate that a slightly positive relationship exists between feed price and eggs/100 layers. This is counter intuitive to preconceived theories that a negative relationship may exist.

### Modeling and Results

Data for this analysis were compiled from multiple sources, including
USDA egg production reports, agricultural price data, and NOAA climate
records. 25 of the most prolific egg-producing states were selected for
study and monthly observations from 2000 through 2024 were merged in
order to create a unified set. Feed price was constructed as a weighted
combination of corn and soybean prices, defined as $\frac{2}{3}$ corn
price and $\frac{1}{3}$ soybean price, to reflect approximate poultry
feed composition. Missing values were substituted with "NA".

Model specification and implementation were derived from recommendations
for linear mixed-effects modeling as outlined by An Introduction to
Linear Mixed-Effects Modeling in R (@brown2021introduction). These
guidelines highlight the importance of matching model structure to the
hierarchical nature of the data, selecting an appropriate random-effects
structure, and prioritizing readability over unnecessary model
complexity. Accordingly, a random-intercept model was utilized to
capture between-state variability while maintaining model stability and
facilitating clear interpretation of fixed effects.

------------------------------------------------------------------------

<div style="text-align: center;">
**Model 1**
</div>
$$
y =
\beta_0
+
\beta_1 \text{avgTemp}
+
\beta_2 \text{feedPrice}
+
b_{0} \text{betweenStateEffects}
+
\varepsilon \text{withinStateEffects}
$$

<div style="text-align: center;">
**Model 2**
</div>
$$
y =
\beta_0
+
\beta_3 \text{Year}
+
b_{0} \text{betweenStateEffects}
+
\varepsilon \text{withinStateEffects}
$$
<div style="text-align: center;">
**Model 3**
</div>
$$
y =
\beta_0
+
\sum_{m=2}^{12} \beta_4 \text{Month}
+
b_{0} \text{betweenStateEffects}
+
\varepsilon \text{withinStateEffects}
$$ 
Model 1 examines egg production as a function of average monthly temperature and feed price, capturing key environmental and economic influences on productivity. Model 2 focuses on long-term trends, modeling changes in egg production from 2000 to 2024. Model 3 evaluates seasonal patterns by incorporating month as a categorical predictor, allowing for the identification of recurring fluctuations in productivity throughout the year. In all models, a random intercept for state is included to account for differences in baseline egg production across states. The residual error term represents within-state variation not explained by the included predictors or random effects.

```{r, warning=FALSE, echo=TRUE}
library(lme4)
library(lmerTest)
library(tidyverse)
library(broom.mixed)
library(gtsummary)

eggData_clean$Month_label <- factor(
  eggData_clean$Month,
  levels = 1:12,
  labels = month.abb
  )

```


```{r, warning=FALSE, echo=TRUE}

#modeling temp and feed price
egg_mod1 <- lmer(
  eggsPer100Layers ~ avgTemp + feedPrice + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod1,
  conf.int = TRUE,
  label = list(
    avgTemp ~ "Average Temperature",
    feedPrice ~ "Feed Price"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 1: Linear Mixed Model Results for Temperature and Feed Price**")
```

```{r, warning=FALSE, echo=TRUE}
#modeling year
egg_mod2 <- lmer(
  eggsPer100Layers ~  Year + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod2,
  conf.int = TRUE,
  label = list(
    Year ~ "Year"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 2: Linear Mixed Model Results for Year**")
```

```{r, warning=FALSE, echo=TRUE}


#modeling months
egg_mod3 <- lmer(
  eggsPer100Layers ~ Month_label + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod3,
  conf.int = TRUE
) %>%
  bold_labels() %>%
  modify_caption("**Table 3: Linear Mixed Model Results for Month**")
```

------------------------------------------------------------------------

The results from the mixed-effects models highlight several key factors influencing egg production. In Model 1, both average temperature and feed price were found to be statistically significant predictors of egg production, with feed price exhibiting a substantially larger effect size. This suggests that economic conditions may play a stronger role in shaping productivity than environmental factors alone.

Model 2 revealed a strong, positive temporal trend, with egg production increasing by approximately 12 eggs per 100 layers per year. This may be indicative of substantial improvements in production efficiency over the study period.

Model 3 demonstrated clear seasonal variation in egg production, with several months showing significant deviations from the reference month. Notably, production tends to decline in late winter and early spring and increase in late summer and winter months, reflecting recurring cyclical patterns in productivity.

Together, these findings indicate that egg production is driven by a combination of long-term industry improvements, seasonal fluctuations, and external environmental and economic factors.

```{r, warning=FALSE, echo=TRUE}
model1_data <- model.frame(egg_mod1)

model1_data$fitted <- fitted(egg_mod1)

ggplot(model1_data, aes(x = fitted, y = eggsPer100Layers)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = "Observed vs Fitted Values (Model 1)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 5"
  ) +
  theme_minimal()
```
Figure 5 indicates that Model 1 reasonably captures the overall linear trend in egg production, as evidenced by the alignment of points along the red reference line. However, there is considerable cloud of observation points around the line, suggesting that much variability remains unexplained. From this, it can be deduced that while the model provides a baseline fit, further refinement, such as incorporating seasonal effects, may improve prediction accuracy.

```{r, warning=FALSE, echo=TRUE}
library(ggeffects)

plot(ggpredict(egg_mod1, terms = "avgTemp")) +
  labs(
    title = "Effect of Temperature on Egg Production (Model 1)",
    x = "Average Monthly Temperature (°F)",
    y = "Predicted Eggs per 100 Layers",
    caption = "Figure 6"
  )

plot(ggpredict(egg_mod1, terms = "feedPrice")) +
  labs(
    title = "Effect of Feed Price on Egg Production (Model 1)",
    x = "Feed Price (USD/bushel)",
    y = "Predicted Eggs per 100 Layers",
    caption = "Figure 7"
  )
```



------------------------------------------------------------------------

```{r, warning=FALSE, echo=TRUE}
model2_data <- model.frame(egg_mod2)

model2_data$fitted <- fitted(egg_mod2)

ggplot(model2_data, aes(x = fitted, y = eggsPer100Layers)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = "Observed vs Fitted Values (Model 2)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 8"
  ) +
  theme_minimal()
```


```{r, warning=FALSE, echo=TRUE}
model2_data <- model.frame(egg_mod2)

model2_data$fitted_year <- fitted(egg_mod2)

ggplot(model2_data, aes(x = Year)) +
  geom_point(aes(y = eggsPer100Layers), alpha = 0.1) +
  geom_line(aes(y = fitted_year), color = "red", linewidth = 1) +
  labs(
    title = "Egg Production Over Time (Model 2)",
    x = "Year",
    y = "Egg Production per 100 Layers",
    caption = "Figure 9"
  )
```
Figure 9 displays observed egg production over time along with fitted values from the linear mixed-effects model. The scatter of points represents individual observations, while the red line shows the model-estimated trend. Despite considerable variability in the observed data, the model captures the overall upward trajectory, suggesting that year is a strong predictor of long-term changes in productivity.

```{r, warning=FALSE, echo=TRUE}
model3_data <- model.frame(egg_mod3)

model3_data$fitted <- fitted(egg_mod3)

ggplot(model3_data, aes(x = fitted, y = eggsPer100Layers)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = "Observed vs Fitted Values (Model 3)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 10"
  ) +
  theme_minimal()
```


```{r, warning=FALSE, echo=TRUE}
eggData_clean %>%
  group_by(Month) %>%
  summarise(mean_eggs = mean(eggsPer100Layers, na.rm = TRUE)) %>%
  ggplot(aes(x = Month, y = mean_eggs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Egg Production Per Month (Model 3)",
    x = "Month",
    y = "Egg Production per 100 Layers",
    caption = "Figure 11"
  )
```


```{r}

```


------------------------------------------------------------------------

```{r, warning=FALSE, echo=TRUE}
re_state <- ranef(egg_mod1)$state
re_state$state <- rownames(re_state)

ggplot(re_state, aes(x = reorder(state, `(Intercept)`), y = `(Intercept)`)) +
  geom_point() +
  coord_flip() +
  labs(
    title = "State-Level Random Intercepts",
    x = "State",
    y = "Deviation from Overall Mean",
    caption = "Figure 12")
```

Figure 12 represents each state’s deviation from the overall mean egg
production. States with positive values exhibit higher baseline
productivity, while those with negative values fall below the overall
average. The wide spread of values indicates substantial variability in egg production across states and highlights the importance of including state-level random effects in the model.

### Conclusion

This analysis examined the relationship between environmental, economic,
and chronological factors and egg production using a linear
mixed-effects model. The results demonstrated a strong upward trend in
egg productivity over time, indicating substantial long-term
improvements in production efficiency across the U.S. egg industry. Feed
price was found to have a positive association with productivity,
suggesting that economic conditions may influence management decisions
and efficiency. By comparison, average temperature exhibited a more mild
effect, indicating that environmental conditions play a smaller but
still measurable role in production outcomes.

In addition to these fixed effects, the model identified significant
variability between states, with state-level differences accounting for
a large proportion of the total variation in egg production. This
finding highlights the importance of regional factors such as management
practices, production systems, and local conditions, which are not fully
captured by the observed predictors.

Overall, the results demonstrate that egg production is driven by a
combination of long-term industry improvements, seasonal patterns, and
substantial differences across states. By incorporating both fixed and
random effects, the mixed-effects modeling approach provides a
comprehensive framework for understanding these dynamics. These findings
suggest that while broad economic and environmental factors influence
egg productivity, state-level characteristics play a critical role in
determining baseline production levels. This has important implications
for both producers and policymakers, as efforts to improve productivity
may benefit from strategies tailored to regional conditions rather than
relying solely on nationwide trends.

## References

ChatGPT-5.2 used to fine tune writing, extract data from original
sources, and develop effective code.

@meteyard2020best @brown2021introduction @zhou2022using
@adhikari2016comparing @ryoo2011model @schielzeth2020robustness
@huitema2025fisher @henderson1953variance @NOAA_ClimateAtAGlance_2026
@USDA_ChickensAndEggs_2026 @USDA_AgriculturalPrices_2026
@Kim2024_HeatStress_LayingHens @Ao2007_MixedModels_HBS
@Mustafa2023_FixedRandomEffects_Medium @sarafian2020lmm
@incredibleegg2013
