---
title: "U.S. Egg Production Analysis Using Linear Mixed Models"
subtitle: "An Exploration Across States and Months"
author: "Rose Dalton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme:
      - sky
      - custom2.scss
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction

<div style="font-size: 0.8em;">

- Egg production varies across time, season, and geography
- Traditional models (ANOVA, regression) assume independence
- This dataset contains:
  - Repeated monthly observations
  - Nested structure (months within states)

→ Linear Mixed Models (LMMs) are appropriate

</div>

## Research Objective

<div style="font-size: 0.8em;">

- Identify key drivers of egg production:
  - Environmental (temperature)
  - Economic (feed price)
  - Temporal (year, month)
- Evaluate:
  - Long-term trends
  - Seasonal patterns
  - State-level variation
  
</div>

## Data Overview

<div style="font-size: 0.8em;">

**Dataset**

- 25 U.S. states
- Monthly data (2000–2024)
- 7,491 observations

**Variables:**

- Response: 
  - Eggs per 100 layers
-Predictors:
  - Temperature: monthly average by state
  - Feed price: USD/bushel
  - Month, Year

</div>

## Why Mixed Models?

<div style="font-size: 0.8em;">

- Observations are **not independent**
- States differ in baseline production

Model structure:
$$
\mathbf{y} = X\boldsymbol{\beta} + Z\boldsymbol{\gamma} + \mathbf{e}
$$

- Fixed effects ($\boldsymbol{\beta}$): temperature, feed, time, season

- Random effects ($\boldsymbol{\gamma}$): state

- Residual (**e**): unexplained variation

</div>


## Exploratory Findings

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# loading packages
library(tidyverse)
library(ggthemes)
library(lme4)
library(lmerTest)
library(tidyverse)
library(broom.mixed)
library(gtsummary)
library(ggeffects)
library(gt)
```

```{r, warning=FALSE, echo=FALSE}
# load data
eggData <- read.csv("USDA_Egg_Data.csv")

# clean data for visuals
eggData_clean <- eggData %>%
  dplyr::select(date, state, eggsPer100Layers, avgTemp, feedPrice) %>%
  mutate(
    state = stringr::str_trim(state),                         
    eggsPer100Layers = as.numeric(eggsPer100Layers), 
    avgTemp = as.numeric(avgTemp),
    Date = as.Date(paste0(date, "-01")),             
    Year = lubridate::year(Date),                              
    Month = lubridate::month(Date)
  )

#time series plot
timeSeries = eggData_clean %>% ggplot(mapping = aes(x=Date, y=eggsPer100Layers)) 

timeSeries + geom_line(color = "#008080", alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, span = 0.25, color = "navy") +
  facet_wrap(~ state, ncol = 5, scales = "free_y") +
  labs(title = "Eggs per 100 Layers Over Time (2000–2024)",
    x = "Date",
    y = "Eggs per 100 Layers",
    caption = "Figure 1") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

<div style="font-size: 0.8em;">

- General upward trend over time
- Variation across states

</div>

## Exploratory Findings

```{r, warning=FALSE, echo=FALSE}
# month box plot
boxPlot = eggData_clean %>% ggplot(mapping = aes(x=factor(Month), y=eggsPer100Layers))

boxPlot +
geom_boxplot(
  fill = "#008080",        
  color = "navy",          
  outlier.alpha = 0.3) +
labs(
  title = "Seasonality: Eggs per 100 Layers by Month",
  x = "Month",
  y = "Eggs per 100 Layers",
  caption = "Figure 2") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

- Production disparities across months

## Exploratory Findings

```{r, warning=FALSE, echo=FALSE}

# temperature scatterplot
scatter = eggData_clean %>% ggplot(mapping = aes(x=avgTemp, y=eggsPer100Layers))

scatter + geom_point(color = "#008080", alpha = 0.25) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75, color = "navy") + 
  labs(title = "Eggs per 100 Layers vs Average Temperature",
    x = "Average Temperature (°F)",
    y = "Eggs per 100 Layers",
    caption = "Figure 3") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

- Temperature shows nonlinear behavior at extremes

## Exploratory Findings

```{r, warning=FALSE, echo=FALSE}

# feed scatterplot
scatter2 = eggData_clean %>% ggplot(mapping = aes(x=feedPrice, y=eggsPer100Layers))

scatter2 + geom_point(color = "#008080", alpha = 0.25) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75, color = "navy") +
  labs(title = "Eggs per 100 Layers vs Feed Price",
       x = "Feed Price (USD/bushel)",
       y = "Eggs per 100 Layers",
    caption = "Figure 4") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

- Slight positive trend in feed price vs productivity

## Model Specification


<div style="font-size: 0.8em;">

**Four-Model Approach**

- **Model 1**: Temperature + Feed Price
- **Model 2**: Long-term trend (Year)
- **Model 3**: Seasonality (Month)
- **Model 4**: Combined model
- All models include: random intercept (State)

</div>

## Model 1 Results

```{r, warning=FALSE, echo=FALSE}

#modeling temp and feed price
egg_mod1 <- lmer(
  eggsPer100Layers ~ avgTemp + feedPrice + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod1,
  conf.int = TRUE,
  label = list(
    avgTemp ~ "Average Temperature",
    feedPrice ~ "Feed Price"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 1: Linear Mixed Model Results for Temperature and Feed Price**") %>%
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(20)
  )
```

<div style="font-size: 0.8em;">

- Temperature: small positive effect
- Feed price: strong positive effect

→ Large unexplained variability

→ No time or seasonal structure

</div>

## Model 2 Results

```{r, warning=FALSE, echo=FALSE}
#modeling year
egg_mod2 <- lmer(
  eggsPer100Layers ~  Year + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod2,
  conf.int = TRUE,
  label = list(
    Year ~ "Year"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 2: Linear Mixed Model Results for Year**") %>%
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(20)
  )
```

<div style="font-size: 0.8em;">

- Year is highly significant
- +12 eggs per year

→ Captures long-term improvement

→ Misses seasonal variation

</div>

## Model 3 Results

:::: {.columns}

::: {.column width="70%"}

```{r, warning=FALSE, echo=FALSE}
eggData_clean$Month_label <- factor(
  eggData_clean$Month,
  levels = 1:12,
  labels = month.abb
)

egg_mod3 <- lmer(
  eggsPer100Layers ~ Month_label + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod3,
  conf.int = TRUE,
  label = list(
    Month_label ~ "Month"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 3: Linear Mixed Model Results for Month**") %>%
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(20),
    data_row.padding = gt::px(4)   
  )
```

:::

::: {.column width="30%"}

<div style="font-size: 0.8em;">

- Significant monthly effects
- February lowest production, peaks in summer and winter

→ Captures seasonal cycles

→ Still missing long-term + external factors

</div>

:::

::::

## Model 4 Results

:::: {.columns}

::: {.column width="70%"}


```{r, warning=FALSE, echo=FALSE}
# combined model
egg_mod4 <- lmer(
  eggsPer100Layers ~ avgTemp + feedPrice + Year + Month_label + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod4,
  conf.int = TRUE,
  label = list(
    avgTemp ~ "Average Temperature",
    feedPrice ~ "Feed Price",
    Year ~ "Year",
    Month_label ~ "Month"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 4: Linear Mixed Model Results for Combined Model**") %>%
  as_gt() %>%
  gt::tab_options(
    data_row.padding = gt::px(3),   # ↓ reduce row height
    table.font.size = gt::px(18)    # ↓ slightly smaller text
  )
```

:::

::: {.column width="30%"}

<div style="font-size: 0.8em;">

- Includes:
  - Temperature
  - Feed price
  - Year
  - Month

→ Most complete representation of production dynamics

</div>

:::

::::

## Model Comparison (AIC)

```{r, warning=FALSE, echo=FALSE}
aic_results <- AIC(egg_mod1, egg_mod2, egg_mod3, egg_mod4) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Model") %>%
  arrange(AIC)
aic_results %>%
  gt() %>%
  tab_header(
    title = "Model Comparison Using AIC"
  )
```

<div style="font-size: 0.8em;">

- Model 4: Lowest AIC (best fit)
- Large improvement over all other models

→ Strong evidence for combined model superiority

</div>

## Model 1 Fit: Observed vs Fitted

```{r, warning=FALSE, echo=FALSE}
model1_data <- model.frame(egg_mod1)

model1_data$fitted <- fitted(egg_mod1)

ggplot(model1_data, aes(x = fitted, y = eggsPer100Layers)) +
  geom_point(color = "#008080", alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "navy") +
  labs(
    title = "Observed vs Fitted Values (Model 1)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 5"
  ) +
  theme_minimal() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

## Model 2 Fit: Observed vs Fitted

```{r, warning=FALSE, echo=FALSE}
model2_data <- model.frame(egg_mod2)

model2_data$fitted <- fitted(egg_mod2)

ggplot(model2_data, aes(x = fitted, y = eggsPer100Layers)) +
  geom_point(color = "#008080", alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "navy") +
  labs(
    title = "Observed vs Fitted Values (Model 2)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 9"
  ) +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

## Model 3 Fit: Observed vs Fitted

```{r, warning=FALSE, echo=FALSE}
model3_data <- model.frame(egg_mod3)

model3_data$fitted <- fitted(egg_mod3)

ggplot(model3_data, aes(x = fitted, y = eggsPer100Layers)) +
  geom_point(color = "#008080", alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "navy") +
  labs(
    title = "Observed vs Fitted Values (Model 3)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 11"
  ) +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

## Model 4 Fit: Observed vs Fitted

```{r, warning=FALSE, echo=FALSE}
full_model_data <- model.frame(egg_mod4)

full_model_data$observed <- model.response(full_model_data)
full_model_data$fitted <- fitted(egg_mod4)

ggplot(full_model_data, aes(x = fitted, y = observed)) +
  geom_point(color = "#008080", alpha = 0.25) +
  geom_abline(slope = 1, intercept = 0, color = "navy", linewidth = 1) +
  labs(
    title = "Observed vs Fitted Values (Combined Model)",
    x = "Fitted Values",
    y = "Observed Eggs per 100 Layers",
    caption = "Figure 14"
  ) +
  theme_minimal()
```

<div style="font-size: 0.8em;">

- Tight clustering around reference line
- Improved predictive accuracy

→ Model 4 explains substantially more variability

</div>

## Model 1 Fit: Residuals vs Fitted

```{r, warning=FALSE, echo=FALSE}
model1_data$residuals <- resid(egg_mod1)

ggplot(model1_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "#008080", alpha = 0.25) +
  geom_hline(yintercept = 0, color = "navy", linewidth = 1) +
  labs(
    title = "Residuals vs Fitted Values (Model 1)",
    x = "Fitted Values",
    y = "Residuals",
    caption = "Figure 6"
  ) +
  theme_minimal()
```

## Model 2 Fit: Residuals vs Fitted

```{r, warning=FALSE, echo=FALSE}
model2_data$residuals <- resid(egg_mod2)

ggplot(model2_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "#008080", alpha = 0.25) +
  geom_hline(yintercept = 0, color = "navy", linewidth = 1) +
  labs(
    title = "Residuals vs Fitted Values (Model 2)",
    x = "Fitted Values",
    y = "Residuals",
    caption = "Figure 10"
  ) +
  theme_minimal()
```

## Model 3 Fit: Residuals vs Fitted

```{r, warning=FALSE, echo=FALSE}
model3_data$residuals <- resid(egg_mod3)

ggplot(model3_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "#008080", alpha = 0.25) +
  geom_hline(yintercept = 0, color = "navy", linewidth = 1) +
  labs(
    title = "Residuals vs Fitted Values (Model 3)",
    x = "Fitted Values",
    y = "Residuals",
    caption = "Figure 12"
  ) +
  theme_minimal()
```

## Model 4 Fit: Residuals vs Fitted

```{r, warning=FALSE, echo=FALSE}
full_model_data$residuals <- resid(egg_mod4)

ggplot(full_model_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "#008080", alpha = 0.25) +
  geom_hline(yintercept = 0, color = "navy", linewidth = 1) +
  labs(
    title = "Residuals vs Fitted Values (Combined Model)",
    x = "Fitted Values",
    y = "Residuals",
    caption = "Figure 15"
  ) +
  theme_minimal()
```

<div style="font-size: 0.8em;">

- Residuals centered around zero
- Reduced heteroskedasticity
- No strong nonlinear patterns

→ Model assumptions reasonably satisfied

</div>

## Random Effects

```{r, warning=FALSE, echo=FALSE}
re_state <- ranef(egg_mod1)$state
re_state$state <- rownames(re_state)

ggplot(re_state, aes(x = reorder(state, `(Intercept)`), y = `(Intercept)`)) +
  geom_point(color = "#008080") +
  coord_flip() +
  labs(
    title = "State-Level Random Intercepts",
    x = "State",
    y = "Deviation from Overall Mean",
    caption = "Figure 11") + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

<div style="font-size: 0.8em;">

- Large spread across states
- Persistent baseline differences

→ Unobserved structural factors matter

</div>

## Key Findings


<div style="font-size: 0.8em;">

- Year is the strongest driver
- Seasonality plays a major role
- Temperature has a modest effect
- Feed price effect weakens when controlling for time

</div>

## Interpretation


<div style="font-size: 0.8em;">

- Production improvements driven by:
  - Technology
  - Efficiency gains
- Seasonal fluctuations:
  - Biological cycles
  - Environmental conditions
- State differences:
  - Management practices
  - Regional conditions

</div>

## Limitations

<div style="font-size: 0.8em;">

- Linear assumption may miss nonlinear effects
- No explicit modeling of:
  - Farm practices
  - Breed differences
- Some heteroskedasticity remains

</div>

## Conclusion

<div style="font-size: 0.8em;">

- Egg production is driven by:
  - Seasonal cycles
  - Long-term industry improvements
  - State-level differences
  
- Simpler models can produce misleading conclusions

- The combined model provides the most reliable and complete representation

- Accounting for hierarchical structure is essential in real-world data


</div>

**→ Mixed models provide a powerful framework for capturing these dynamics.**

## Thank You

<div style="display: flex; justify-content: left; align-items: center; height: 20vh;">

Questions?

</div>



