---
title: "U.S. Egg Production Analysis Using Linear Mixed Models"
subtitle: "An Exploration Across States and Months"
author: "Rose Dalton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme:
      - sky
      - custom2.scss
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Motivation

**Why Study Egg Production?**

- Productivity varies across:
  - Time
  - Season
  - State
  
- Traditional models assume independence

- Data are hierarchical + repeated

**Goal:** Identify key drivers of productivity

## Why Linear Mixed Models?

**Nature of the Data**

- Observations nested within states
- Repeated measurements over time

**Mixed model allows:**

- Fixed effects → global trends
- Random effects → state variation

## Data Exploration and Visualization

**Dataset**

- 25 U.S. states
- Monthly data (2000–2024)
- 7,491 observations

**Variables:**

- Eggs per 100 layers
- Temperature: monthly average by state
- Feed price: USD/bushel
- Month, Year


## Model Strategy

**Three-Model Approach**

- Model 1:

  - Temperature + Feed Price

- Model 2:

  - Long-term trend (Year)

- Model 3:

  - Seasonality (Month)

- All models include:

  - Random intercept (State)

## Model Framework

**Linear Mixed Model**
$$
\mathbf{y} = X\boldsymbol{\beta} + Z\boldsymbol{\gamma} + \mathbf{e}
$$

- Fixed effects: temperature, feed, time, season

- Random effects: state

- Residual: unexplained variation

## Data Exploration and Visualization {.smaller}

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(ggthemes)
library(lme4)
library(lmerTest)
library(tidyverse)
library(broom.mixed)
library(gtsummary)
library(ggeffects)
```

```{r, warning=FALSE, echo=F}
# load data
eggData <- read.csv("USDA_Egg_Data.csv")

# clean data for visuals
eggData_clean <- eggData %>%
  select(date, state, eggsPer100Layers, avgTemp, feedPrice) %>%
  mutate(
    state = str_trim(state),                         
    eggsPer100Layers = as.numeric(eggsPer100Layers), 
    avgTemp = as.numeric(avgTemp),
    Date = as.Date(paste0(date, "-01")),             
    Year = year(Date),                              
    Month = month(Date)
  ) 

#time series plot
timeSeries = eggData_clean %>% ggplot(mapping = aes(x=Date, y=eggsPer100Layers)) 

timeSeries + geom_line(color = "#008080", alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, span = 0.25, color = "navy") +
  facet_wrap(~ state, ncol = 5, scales = "free_y") +
  labs(title = "Eggs per 100 Layers Over Time (2000–2024)",
    x = "Date",
    y = "Eggs per 100 Layers",
    caption = "Figure 1") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```
→ Much state-level variation egg production

## Data Exploration and Visualization {.smaller}

```{r, warning=FALSE, echo=FALSE}
# month box plot
boxPlot = eggData_clean %>% ggplot(mapping = aes(x=factor(Month), y=eggsPer100Layers))

boxPlot +
geom_boxplot(
  fill = "#008080",        
  color = "navy",          
  outlier.alpha = 0.3) +
labs(
  title = "Seasonality: Eggs per 100 Layers by Month",
  x = "Month",
  y = "Eggs per 100 Layers",
  caption = "Figure 2") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```
→ Some monthly-level variation

## Data Exploration and Visualization {.smaller}

```{r, warning=FALSE, echo=FALSE}

# temperature scatterplot
scatter = eggData_clean %>% ggplot(mapping = aes(x=avgTemp, y=eggsPer100Layers))

scatter + geom_point(color = "#008080", alpha = 0.25) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75, color = "navy") + 
  labs(title = "Eggs per 100 Layers vs Average Temperature",
    x = "Average Temperature (°F)",
    y = "Eggs per 100 Layers",
    caption = "Figure 3") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```
→ Warmer weather = sustained productivity, until a certain threshold

## Data Exploration and Visualization {.smaller}

```{r, warning=FALSE, echo=FALSE}

# feed scatterplot
scatter2 = eggData_clean %>% ggplot(mapping = aes(x=feedPrice, y=eggsPer100Layers))

scatter2 + geom_point(color = "#008080", alpha = 0.25) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75, color = "navy") +
  labs(title = "Eggs per 100 Layers vs Feed Price",
       x = "Feed Price (USD/bushel)",
       y = "Eggs per 100 Layers",
    caption = "Figure 4") +
  theme_bw() + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```
→ Indicates slight positive trend

## Model 1 Results

```{r, warning=FALSE, echo=FALSE}

#modeling temp and feed price
egg_mod1 <- lmer(
  eggsPer100Layers ~ avgTemp + feedPrice + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod1,
  conf.int = TRUE,
  label = list(
    avgTemp ~ "Average Temperature",
    feedPrice ~ "Feed Price"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 1: Linear Mixed Model Results for Temperature and Feed Price**") %>%
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(20)
  )
```

- Both statistically significant
- Feed price has larger effect
- Wide uncertainty for temperature

→ Likely reflects time/economic trends, not causal

## Model 2 Results

```{r, warning=FALSE, echo=FALSE}
#modeling year
egg_mod2 <- lmer(
  eggsPer100Layers ~  Year + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod2,
  conf.int = TRUE,
  label = list(
    Year ~ "Year"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 2: Linear Mixed Model Results for Year**") %>%
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(20)
  )
```

- +12 eggs per year
- Strong upward trajectory

→ Likely industry efficiency improvements

## Model 3 Results

:::: {.columns}

::: {.column width="70%"}

```{r, warning=FALSE, echo=FALSE}
eggData_clean$Month_label <- factor(
  eggData_clean$Month,
  levels = 1:12,
  labels = month.abb
)

egg_mod3 <- lmer(
  eggsPer100Layers ~ Month_label + (1 | state),
  data = eggData_clean
)

tbl_regression(
  egg_mod3,
  conf.int = TRUE,
  label = list(
    Month_label ~ "Month"
  )
) %>%
  bold_labels() %>%
  modify_caption("**Table 3: Linear Mixed Model Results for Month**") %>%
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(14)
  )
```

:::

::: {.column width="30%"}

- Strong cyclical pattern → Seasonality explains large variation

:::

::::

## Model Comparison

**Which Model Performs Best?**

- Model 1 → baseline fit
- Model 2 → improved (captures trend)
- Model 3 → best fit
  - Reduced residual variability
  - Captures cyclical structure

## Random Effects

```{r, warning=FALSE, echo=FALSE}
re_state <- ranef(egg_mod1)$state
re_state$state <- rownames(re_state)

ggplot(re_state, aes(x = reorder(state, `(Intercept)`), y = `(Intercept)`)) +
  geom_point(color = "#008080") +
  coord_flip() +
  labs(
    title = "State-Level Random Intercepts",
    x = "State",
    y = "Deviation from Overall Mean",
    caption = "Figure 11") + 
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5)
)
```

- Large spread across states
- Persistent baseline differences

→ Unobserved structural factors matter

## Key Takeaways

- Strong seasonal pattern (dominant driver)
- Significant long-term growth
- Temperature → modest effect
- Feed price → likely confounded
- Large state variability

## Conclusion

- Egg production is driven by:
  - Seasonal cycles
  - Long-term industry improvements
  - State-level differences

**Mixed models provide a powerful framework for capturing these dynamics.**

## Future Efforts

- Nonlinear temperature effects
- Random slopes by state
- Interaction terms
- Forecasting models


## References
